---
title: "Appendix III - Analyse the Data"
author: "Jamie Cash"
date: "2023-07-06"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
```

```{r load packages, include=FALSE}
# For plotting on UK map
library("cowplot")
library("googleway")
library("ggrepel")
library("ggspatial")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")

library("tidyverse")

# For rolling average against dataset with NAs
library("zoo")
```

# Load the processed data

Weather station data provided by the MET office is provided on their website in txt files. This data was [prepared](prepare.html) and [processed](process.html) to create two summary datasets: [seasonal temperatures by year](seasonal_summary.csv); and [annual temperatures by station](annual_station_summary.csv). Load these datasets.

```{r load seasonal, errors=FALSE}
seasonal <- read.csv('seasonal_summary.csv')
head(seasonal)
```
```{r load station, errors=FALSE}
station <- read.csv('annual_station_summary.csv')
head(station)
```
# Add calculations

Add mean and rolling mean to station data.

```{r add rolling means}
# Calculate means, rolling means and deltas.  We only need 2022 after calculation so 
# filter.
station_means_2022 <- station %>% 
  group_by(station_name) %>% 
  arrange(year) %>% 
  mutate(
      high_mean=mean(high, na.rm=TRUE),
      low_mean=mean(low, na.rm=TRUE),
      rainfall_mean=mean(total_rainfall, na.rm=TRUE),
      high_5yr_roll=rollapplyr(high, 5, partial=TRUE, fill=NA, FUN=function(x) mean(x, na.rm=TRUE)),
      low_5yr_roll=rollapplyr(low, 5, partial=TRUE, fill=NA, FUN=function(x) mean(x, na.rm=TRUE)),
      rainfall_5yr_roll=rollapplyr(total_rainfall, 5, partial=TRUE, fill=NA, FUN=function(x) mean(x, na.rm=TRUE)),
      high_delta=high_5yr_roll-high_mean,
      low_delta=low_5yr_roll-low_mean,
      rainfall_delta=rainfall_5yr_roll-rainfall_mean
    ) %>% 
  filter(year==2022)

head(station_means_2022) 
```

# Create map to plot data against

```{r create UK map}
world <- ne_countries(scale = "medium", returnclass = "sf")

uk_map <- ggplot(data = world) +
    geom_sf() +
    xlab("Longitude") + 
    ylab("Latitude") +
    coord_sf(xlim = c(2, -11), ylim = c(49, 59), expand = FALSE)

uk_map
```

# Answer the questions

## Have the seasons shifted?

* Is the average highest temperature for December, January and February across all UK locations for the last 5 years higher than the  historical average?

### Historical mean of winter highs

```{r winter highs - mean}
# Get mean
seasonal %>% 
  filter(season == 'Winter') %>% 
  summarise(mean_high = mean(high))
```

### 5 year mean of winter highs

```{r winter highs - 5 yr mean}
# Get 5 year mean. Exclude 2023 as we don't have December 2023 data.
seasonal %>% 
  filter(season == 'Winter') %>% 
  filter(between(year, 2018, 2022 )) %>% 
  summarise(mean_high = mean(high))
```

### Chart

```{r winter highs - chart}
plot <- seasonal %>%
  filter(season=="Winter") %>% 
  ggplot(aes(x=year, y=high)) +
  geom_point(color='darkgray') +
  geom_smooth() +
  labs(title="Winter Highs by Year", x="Year", y="Temperature (c)") +
  geom_text(x = 1900, y = 9.86, label = "Mean=9.86") + 
  geom_text(x = 2000, y = 11.54, label = "2018-2022 Mean = 11.54")

ggsave("winter_highs.png", plot)

plot
```

* Is the average lowest temperature for March, April and May across all UK locations for the last 5 years lower than the  historical average?

### Historical mean of spring lows

```{r spring lows - mean}
# Get mean
seasonal %>% 
  filter(season == 'Spring') %>% 
  summarise(mean_low = mean(low))
```

### 5 year mean of spring lows

```{r spring lows - 5 yr mean}
# Get 5 year mean. Exclude 2023 as we don't have December 2023 data.
seasonal %>% 
  filter(season == 'Spring') %>% 
  filter(between(year, 2018, 2022 )) %>% 
  summarise(mean_low = mean(low))
```

### Chart

```{r spring lows - chart}
plot <- seasonal %>%
  filter(season=="Spring") %>% 
  ggplot(aes(x=year, y=low)) +
  geom_point(color='darkgray') +
  geom_smooth() +
  labs(title="Spring Lows by Year", x="Year", y="Temperature (c)") +
  geom_text(x = 1900, y = -0.04, label = "Mean=-0.04") + 
  geom_text(x = 2000, y = -1.2, label = "2018-2022 Mean=-1.2")

ggsave("spring_lows.png", plot)

plot
```

## Are we gettig higher temperature peaks?

* For each location, is the average highest annual temperature over the last 5 years higher than the historical average?

Get the min and max values for the scale.

```{r min and max high delta}
min(station_means_2022$high_delta)
max(station_means_2022$high_delta)
```
Plot the deltas on the chart. Scale should be even. We will use -2 to 2 as 1.9 
is the max.

```{r temperature high deltas map}
high_delta_map <- uk_map +
  geom_point(data = station_means_2022, 
             mapping = aes(x=station_long, y=station_lat, colour=high_delta, 
                           size=high_delta)) +
  scale_colour_gradient(low="blue", high="red", limits = c(-2,2)) +
  labs(title="High Temperature Change", 
       subtitle="Delta between average 5 year high and average high since 1978")

ggsave("station_high_deltas.png", high_delta_map)

high_delta_map
```

## Are we getting lower temperature dips?

* For each location, is the average lowest annual temperature over the last 5 years lower than the historical average?

Plot the deltas on the chart. Scale should the same as the previous chart for consistency.

Get the min and max values for the report.

```{r min and max low delta}
min(station_means_2022$low_delta)
max(station_means_2022$low_delta)
```
Some stations have a lower low. Which ones? We will include these in the report.

```{r stations with lower low}
station_means_2022 %>% filter(low_delta <=0)
```

```{r temperature low deltas map}
low_delta_map <- uk_map +
  geom_point(data = station_means_2022, 
             mapping = aes(x=station_long, y=station_lat, colour=low_delta, 
                           size=low_delta)) +
  scale_colour_gradient(low="blue", high="red", limits = c(-2,2)) +
  labs(title="Low Temperature Change", 
       subtitle="Delta between average 5 year low and average low since 1978")

ggsave("station_low_deltas.png", low_delta_map)

low_delta_map
```

## Are we getting more or less rainfall?

* For each location, is the average yearly rainfall for the last 5 years lower or higher than the historical average?

Get the min and max values for the scale.

```{r min and max rainfall delta}
min(station_means_2022$rainfall_delta)
max(station_means_2022$rainfall_delta)
```

Plot the deltas on the chart. Scale should be even. We will use -300 to 300 as highest delta is 292.

```{r rainfall deltas map}
rainfall_delta_map <- uk_map +
  geom_point(data = station_means_2022, mapping = 
               aes(x=station_long, y=station_lat, colour=rainfall_delta, 
                   size=rainfall_delta)) +
  scale_colour_gradient(low="brown", high="green", limits = c(-300,300)) +
  labs(title="Rainfall Change", 
       subtitle="Delta between average 5 year rainfall and average high since 1978")

ggsave("station_rainfall_deltas.png", rainfall_delta_map)

rainfall_delta_map
```

