---
title: "Appendix II - Process the Data"
author: "Jamie Cash"
date: "2023-07-05"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
```

```{r load packages, include=FALSE}
library("tidyverse")
```

# Load the prepared data

Weather station data provided by the MET office is provided on their website in txt files. This data was [prepared](prepare.html) to create a clean dataset which is available [here](met_station_readings.csv). Load this dataset.

```{r load data, errors=FALSE}
data <- read.csv('met_station_readings.csv', colClasses = 
                   c("character", "Date", "numeric", "numeric", "numeric"))
head(data)
```

# Process the data

## Seasonal temperatures by year
Summarise the data to create a dataset containing seasonal high, low, average max and average min temperature across all weather stations by year.

```{r seasonal data, errors=FALSE}
# Add season and year columns
data <- data %>% 
  mutate(season = 
    case_when(
      month(year_month) %in% c(12, 1, 2) ~ 'Winter',
      month(year_month) %in% c(3, 4, 5) ~ 'Spring',
      month(year_month) %in% c(6, 7, 8) ~ 'Summer',
      month(year_month) %in% c(9, 10, 11) ~ 'Autumn'
    )
  )

data$year = year(data$year_month)

# Group data by season and compute summary columns
seasonal <- data %>%
  group_by(year, season) %>%
  summarise(high = max(max_temp), low = min(min_temp), average_max = mean(max_temp), average_min = mean(min_temp)) %>%
  arrange(year, season)

# Drop NA rows
seasonal <- na.omit(seasonal)

# Save
write.csv(seasonal, "seasonal.csv", row.names=FALSE)
head(seasonal)
```
## Seasonal temperatures by year
Summarise the data to create a dataset containing annual high, low and highest monthly rainfall by station by year.

```{r annual station data, errors=FALSE}
# Group data by year and station and compute summary columns
annual_station <- data %>%
  group_by(year, station_name) %>%
  summarise(high = max(max_temp), low = min(min_temp), max_monthly_rainfall = max(rainfall)) %>%
  arrange(year, station_name)

# Save
write.csv(annual_station, "annual_station.csv", row.names=FALSE)
head(annual_station)
```